---
title: Are different marital statuses and family income correlated in the United States
  of America?
output:
  html_document:
    theme: cerulean
  pdf_document: default
date: '2014-10-15'
---

<!-- For more info on RMarkdown see http://rmarkdown.rstudio.com/ -->

<!-- Enter the code required to load your data in the space below. The data will be loaded but the line of code won't show up in your write up (echo=FALSE) in order to save space-->
```{r echo=FALSE}
options(width=160, scipen=100, digits=3)
#load(url("http://bit.ly/dasi_gss_data"))
load("statistics_project_gss.Rdata")
```

<!-- In the remainder of the document, add R code chunks as needed -->

### Introduction:

The research question for this analysis looks at the correlation between martial status (Married and Not Married) and family income within the United States of America. In this analysis we are looking at the mean of family income to establish correlation across different marital statuses. A common assumption is that a stable marriage has economic benefits for both partners. Another common assumption is that more high income adults have a higher marriage rate. A correlation would be an indicator that either of those assumption could be true even if causality cannot be established.

### Data:

The data for this analysis comes from the General Social Survey (GSS), derived from in-person interviews conducted by the National Opinion Research Center at the University of Chicago, of adult residents of the USA. Each case represents the answers of a randomly selected adult. There are 57061 cases in the dataset.

The variables of interest from the data set are "coninc", a numerical continuous variable representing the family income in constant dollars, and "marital", a categorical (not ordinal) variable representing the marital status with values Married, Widowed, Divorced, Separated, or Never Married. We are combining the statuses Widowed, Divorced, Separated and Never Married into a new status called Not Married to facilitate hypotheses testing for this analysis. All NA values for "marital" and "coninc" have been removed from the data as well.

This is an observational study as the data is from a survey of adults selected from families across the whole United States. Therefore the population of interest are the residents of the United States of America. The study is based on stratified random sampling collected every other year from 1972 to 2012. The strata changed over time but geographical strata were usually included. Special effort was taken to avoid non-responsive bias by calling selected participants outside of working hours and on weekends. There is still the possibility of selection bias as the sample size is small with a mean of 2853 cases per year and participants are required to participate in an in-person interview. In general the sample can be considered generalizable. As the data comes from an observational study causality cannot be established between the variables only correlation.

### Exploratory data analysis:

```{r echo=FALSE}
gssredux <- subset(gss, gss$coninc !='NA' & gss$marital !='NA')
gssredux <- gssredux[c("coninc", "marital")]

marital_merged <- as.character(gssredux$marital)
marital_merged[marital_merged %in% c('Widowed', 'Divorced', 'Separated', 'Never Married')] <- "Not Married"
marital_merged <- as.factor(marital_merged)
marital_merged <- droplevels(marital_merged)
gssredux$marital_merged <- marital_merged

married <- gssredux[gssredux$marital_merged == 'Married',]
notmarried <- gssredux[gssredux$marital_merged == 'Not Married',]
```

* Summary of the variable family income in the GSS data set
```{r echo=FALSE}
summary(gssredux$coninc)
```
* Summary of the variable marital status in the GSS data set
```{r echo=FALSE}
summary(gssredux$marital)
```

* Histogram for variable family income

```{r echo=FALSE}
hist(gssredux$coninc, main=NULL, xlab="Family Income")
```

* Histogram for variable combined marital status (married vs not married)

```{r echo=FALSE}
barplot(summary(gssredux$marital_merged), xlab="Marital Status", ylab="Frequency")
```

* Boxplot for both variables, family income and combined martial status

```{r echo=FALSE}
plot(gssredux$marital_merged, gssredux$coninc, xlab="Marital Status", ylab="Family Income")
```

Explanatory analysis suggest that there is a large enough amount of cases for each martial status to allow for hypotheses testing. Family income distribution is right skewed and unimodal. The box plot shows a clear difference of the family income between the different martial statuses which might point to a correlation.

### Inference:

#### Hypotheses

__Null Hypotheses__: no difference between the mean of family income of married and not married cases

$H_{0}: \mu_{Married}-\mu_{NotMarried}=0$

__Alternative Hypotheses__: some difference between the mean of family income of married and not married cases 

$H_{A}: \mu_{Married}-\mu_{NotMarried}\neq0$

#### Conditions
* Independence
    * within groups Married and Not Married
        * random sampling is used for all cases in the GSS
        * 51225 cases is less than < 10% of the population 
    * between groups Married and Not Married
        * Married and Not Married are not paired
* Sample size and skew
    * Married is 27931 cases, Not Married is 23294 cases which is each more than the required 30 cases
    * the relatively large sample should also compensate for the skew of family income in the population

#### Inference Method

The confidence interval calculation and hypotheses test will be performed using the central limit theorem for comparing independent means. We can use this method because the response variable (family income) is numeric, the explanatory variable is a categorical variable with only 2 levels (Married and Not Married) and the two groups can be assumed to not be paired.

The point estimate is calculated as a difference between the means, the standard error using the method for independent means $SE_{(\bar{x}_1-\bar{x}_2)}=\sqrt{s_1^2/n_1+s_2^2/n_2}$, followed by the z- and p-values as well as the confidence interval for 95% per standard CLT. If the p value is below the $\alpha$ of 5% we reject the null hypotheses.

#### Perform Inference
```{r}
pe <- mean(married$coninc) - mean(notmarried$coninc)
se = sqrt(sd(married$coninc)^2/length(married$coninc) + sd(notmarried$coninc)^2/length(notmarried$coninc))
z <- (abs(pe) - 0) / se
z95 <- abs(qnorm(0.025))
ci <- c(pe - (z95*se), pe + (z95*se))
p <- 2*pnorm(z, lower.tail = FALSE)
```
Point estimate (difference between means) is `r pe` with a standard error of `r se`. This results in a 95% confidence interval of (`r ci`) and a p-value of `r p`. As p is below 5% we reject the null hypotheses.

#### Interpret Results
As a p-value of `r p` is below the $\alpha$ of 5% we reject the null hypotheses. If there is no difference between the mean family income of married and not married individuals, there is a `r p`% chance of obtaining random samples of `r length(married$coninc)` married and `r length(notmarried$coninc)` not married individuals where the average difference between family income is at least `r pe`$. The 0 difference value of the null hypotheses is also not within (`r ci`), the 95% confidence interval, so both methods agree.

### Conclusion:

Based on the data it was demonstrated that there is a clear difference in average income between individuals who are married and not married in the United States of America. We cannot establish a causation but a clear correlation is shown. A further analysis is needed to control for work status (working or non-working) of subject and subject's spouse for married individuals, to see if double income is strongly influencing the result.
